# Exercise

## import the Tidyverse libraries & import dataset

```{r}
install.packages('tidyverse')

#check if it has been download
library(tidyverse)

```

```{r}
#load the data to the dataframe
a <- read_csv('children_bodysize.csv')
View(a)

#check the data types for each column
names(a)

```

## plot showing heights vs age

```{r}
##challenge
#create a column height containing the stature in cm
a$height <- a$stature/10

#create a column contaning the age in years
a$year_old <- a$age/12

#plot showing heights vs age
plot(height~year_old, data=a)

#plot showing average height of children by age
plot(height~year_old, data=a, pch='.', col='blue')
```

## questions

### 1. how many entries are in the dataset

```{r}
#entries means the row ammount
entry <- nrow(a)
entry
```

### 2. what are the youngest and oldest ages of individuals in the dataset

```{r}
#youngest
young <- min(a$year_old)
young

#oldest
oldest <- max(a$year_old)
oldest
```

### 3. what is the tallest individual in the dataset, are thay male or female?

```{r}
#find the tallest one
tallest <- a[which.max(a$height), ]

#find out if the one is male or female
tallest_gender <- tallest$gender
tallest_gender
```

### 4. how many boys are in the dataset

```{r}
boys <- sum(a$gender == 'M')
boys
```

### 5. how many girls are over 12

```{r}
girl_over_12 <- sum(a$gender == 'F' & a$year_old == 12)
girl_over_12
```

### 6. what is the mean height of a 5 year old boy and a 5 year old girl

```{r}
#5 year old boy
mean_boy <- mean(a$height[a$gender == 'M' & a$year_old == 5])
mean_boy

#5 year old girl
mean_girl <- mean(a$height[a$gender == 'F' & a$year_old == 5])
mean_girl

```

### 7. what is the mean height of a 15 year old boy and 15 year old girl

```{r}
#15 year old boy
mean_boy2 <- mean(a$height[a$gender == 'M' & a$year_old == 15])
mean_boy2

#15 year old girl
mean_girl2 <- mean(a$height[a$gender == 'F' & a$year_old == 15])
mean_girl2
```

### 8. use a histogram to show the distribution of children's age

```{r}
hist(a$year_old, main = 'Distribution of age', xlab = 'age', ylab = 'frequency')
```

### 9. what does the distribution of hte heights of 9 year old males look like

```{r}
plot(height~year_old, data = subset(a, year_old = 9, gender == 'M'))
```
